\pkg{cubble}: An \proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2023-06-21 main.Rmd
update conclusion

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'Universität für Bodenkultur Wiens \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: BNU-HKBU United International College
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolaslangrene@uic.edu.cn
  - name: Patricia Menéndez
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An \\proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Organizing and Wrangling Multivariate Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
     Multivariate spatio-temporal data refers to multiple measurements taken across space and time. For many analyses, spatial and time components can be separately studied: for example, to explore the temporal trend of one variable for a single spatial location, or to model the spatial distribution of one variable at a given time. However for some studies, it is important to analyze different aspects of the spatio-temporal data simultaneously, like for instance, temporal trends of multiple variables across locations. In order to facilitate the study of different portions or combinations of spatio-temporal data, we introduce a new class, \code{cubble}, with a suite of functions enabling easy slicing and dicing on different spatio-temporal components.  The proposed \code{cubble} class ensures that all the components of the data are easy to access and manipulate while providing flexibility for data analysis. In addition, the \pkg{cubble} package facilitates visual and numerical explorations of the data while easing data wrangling and modelling. The \code{cubble} class and the functions provided in the \pkg{cubble} \proglang{R} package equip users with the capability to handle hierarchical spatial and temporal structures. The \code{cubble} class and the tools implemented in the package are illustrated with different examples of Australian climate data.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
output: rticles::jss_article
bibliography: references.bib
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  fig.path = here::here("figures/")
)

@@

This paper presents an \proglang{R} package \pkg{cubble} for organizing, manipulating and visualizing spatio-temporal data. The package introduces a new data class for spatio-temporal data, \code{cubble}, that connects the time invariant and varying variables and that allows the user to work with a nested and long form of the data. This work adds capabilities into the spatio-temporal practitioners toolbox to integrate it with a tidy data framework. The data structure and functions introduced in this package can be used and combined with existing spatial data analysis packages such as \pkg{sf}, data wrangling packages such as \pkg{dplyr}, and visualization packages such as \pkg{ggplot2}, \pkg{plotly} and \pkg{leaflet}.

Numerous examples are provided in the main text, appendix and package vignettes. These include creating and coercing data with mismatched names, handling hierarchical data, matching time series spatially and temporally, reproducing ERA5 plots from NetCDF data. Visualization of the \code{cubble} objects can be done with interactive graphic pipelines using \pkg{crosstalk}, \pkg{plotly} and \pkg{leaflet}.

There are several possible future directions for the work. The data structure only described fixed spatial sites, and it could be useful to provide tools to accommodate moving coordinates as might be encountered in animal movement data. That could be achieved with a list-column for the location coordinates, and an additional form that these locations can be pivoted into, like the long form for temporal variables. For multiple measured variables, the \pkg{cubble} package could be integrated with dimension reduction methods, and dynamic graphics for multiple dimensions such as a tour [@wickham_tourr_2011].

>>>>

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of  Natural `\\`{=latex} Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU `\\`{=latex} United International College'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolaslangrene@uic.edu.cn
  - name: Patricia Menéndez
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An \\proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Organizing and Wrangling Multivariate Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
     Multivariate spatio-temporal data refers to multiple measurements taken across space and time. For many analyses, spatial and time components can be separately studied: for example, to explore the temporal trend of one variable for a single spatial location, or to model the spatial distribution of one variable at a given time. However for some studies, it is important to analyze different aspects of the spatio-temporal data simultaneously, like for instance, temporal trends of multiple variables across locations. In order to facilitate the study of different portions or combinations of spatio-temporal data, we introduce a new class, \code{cubble}, with a suite of functions enabling easy slicing and dicing on different spatio-temporal components.  The proposed \code{cubble} class ensures that all the components of the data are easy to access and manipulate while providing flexibility for data analysis. In addition, the \pkg{cubble} package facilitates visual and numerical explorations of the data while easing data wrangling and modelling. The \code{cubble} class and the functions provided in the \pkg{cubble} \proglang{R} package equip users with the capability to handle hierarchical spatial and temporal structures. The \code{cubble} class and the tools implemented in the package are illustrated with different examples of Australian climate data.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
output: 
  bookdown::pdf_book:
    base_format: rticles::jss_article
bibliography: references.bib
---
<!-- Universität für Bodenkultur Wiens \AND -->
```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  fig.path = here::here("figures/")
)

@@

This paper presents the \proglang{R} package \pkg{cubble} for organizing, wrangling and visualizing spatio-temporal data. The package introduces a new data class, \code{cubble}, consisting of two subclass, spatial cubble and a temporal cubble, to organise spatio-temporal data in two different formats within the tidy data framework. The data structure and functions introduced in this package can be used and combined with existing spatial and temporal data analysis packages such as \pkg{sf} and \pkg{tsibble}, data wrangling packages such as \pkg{dplyr}, and visualization packages such as \pkg{ggplot2}, \pkg{plotly}, and \pkg{leaflet}.

The paper includes numerous examples to illustrate the utility of \pkg{cubble} as a data structure for spatio-temporal analysis. These examples cover different tasks of a typical analysis workflow: handling data with spatial and temporal misalignment, matching data from multiple sources, and creating both static and interactive spatio-temporal visualisation. For future directions, other commonly-used spatial or temporal data structures can be integrated into \pkg{cubble} to extend analysts' familiar spatial and temporal toolkit to spatio-temporal.

### 2023-06-21 main.tex
update conclusion

\author{
H. Sherry Zhang\\Monash University \And Dianne Cook\\Monash
University \And Ursula Laa\\Universität für Bodenkultur Wiens
\AND Nicolas Langrené\\BNU-HKBU United International
College \And Patricia Menéndez\\Monash University
}
\title{\pkg{cubble}: An \proglang{R} Package for Organizing and
Wrangling Multivariate Spatio-temporal Data}

@@

      Ursula Laa\\
    Universität für Bodenkultur Wiens\\
    Gregor-Mendel-Straße 33, 1180 Wien, Austria\\
  E-mail: \href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}\\

      Nicolas Langrené\\
    BNU-HKBU United International College\\
    2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China\\
  E-mail: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}\\

@@

The rest of the paper is organized as follows: Section @ref(cubble)
presents the main design and functionality of the \pkg{cubble} package.
Section @ref(others) explains how the \pkg{cubble} package deals with
more advanced considerations, including data with hierarchical
structure, data matching and how the package fits with existing static
and interactive visualization tools. Moreover we also illustrate how the
\pkg{cubble} package deals with spatio-temporal data transformations.
Section @ref(examples) uses Australian weather station data and river
level data as examples to demonstrate the use of the package. An example
of how the \pkg{cubble} package handles Network Common Data Form
(NetCDF) data is also provided. Section @ref(conclude) discuss the paper
contributions and future directions.

@@

\caption[The full space-time and sparse space-time layouts handled by the cubble package]{The full space-time and sparse space-time layouts handled by the cubble package. Reproduced from Figure 1 in the spacetime paper \citep{spacetime}.}(\#fig:layout)
\end{figure}
\end{CodeChunk}

@@

In a spatial cubble, spaital variables are organised as columns and
temporal variables are nested within a specialised \texttt{ts} column.
The cubble object printed below contains weather records in three
airport stations (the creation of a cubble object will be explained in
Section @ref(create)). This toy data is a subset of a larger data
\code{climate_aus} collected from Global Historical Climatology Network
Daily (GHCND). It records three airport stations located in Melbourne,
Australia and includes spatial variables such as station ID, longitude,
latitude, elevation, station name, World Meteorology Organisation ID.
The dataset contains temporal variables including precipitation, maximum
and minimum temperature, which can be read from the cubble header:

@@

The function \code{make_cubble()} composes a \code{cubble} object from a
spatial table (\code{spatial}) and a temporal table (\code{temporal}),
along with three attributes introduced in the subsection @ref(object):
\code{key}, \code{index}, and \code{coords}. The following code creates
the spatial \code{cubble} object shown previously:

@@

Table @ref(tab:funs) summarises the functions implemented in the
\pkg{cubble} package and Table @ref(tab:methods) details the methods
implemented for each of the three cubble classes. The \code{cubble_df}
class handles methods that behave consistently in both spatial
(\code{spatial_cubble_df}) and temporal cubble
(\code{temporal_cubble_df}). When the method has distinct behavior for
the spatial cubble and temporal cubble, it is implemented separately for
each subclass.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8750}}@{}}
\caption{(\#tab:funs) Functions implemented in the \pkg{cubble}
package}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Category
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Functions
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Category
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Functions
\end{minipage} \\
\midrule()
\endhead
base R & \texttt{{[}}, \texttt{{[}{[}\textless{}-},
\texttt{names\textless{}-} \\
tidyverse & \texttt{dplyr\_row\_slice}, \texttt{dplyr\_col\_modify},
\texttt{dplyr\_reconstruct}, \texttt{select}, \texttt{mutate},
\texttt{arrange}, \texttt{filter}, \texttt{group\_by}, \texttt{ungroup},
\texttt{summarise}, \texttt{select}, \texttt{slice}, \texttt{rowwise},
\texttt{rename}, \texttt{bind\_rows}, \texttt{bind\_cols},
\texttt{relocate}, \texttt{type\_sum}, the slice family
(\texttt{slice\_head}, \texttt{slice\_tail}, \texttt{slice\_max},
\texttt{slice\_min}, \texttt{slice\_sample}) and the join family
(\texttt{left\_join}, \texttt{right\_join}, \texttt{inner\_join},
\texttt{full\_join}, \texttt{anti\_join}, \texttt{semi\_join}) \\
cubble & \texttt{as\_cubble}, \texttt{cubble}, \texttt{make\_cubble},
\texttt{check\_key} \texttt{face\_temporal}, \texttt{face\_spatial},
\texttt{unfold}, \texttt{key}, \texttt{key\_vars}, \texttt{key\_data},
\texttt{index}, \texttt{index\_var}, \texttt{coords}, \texttt{spatial},
\texttt{match\_sites}, \texttt{match\_spatial},
\texttt{match\_temporal}, \texttt{geom\_glyph},
\texttt{geom\_glyph\_box}, \texttt{geom\_glyph\_line},
\texttt{make\_spatial\_sf}, \texttt{make\_temporal\_tsibble},
\texttt{fill\_gaps}, and \texttt{scan\_gaps} \\
\bottomrule()
\end{longtable}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7500}}@{}}
\caption{(\#tab:methods) Methods implemented for the three \code{cubble}
classes.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Class
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Methods
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Class
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Methods
\end{minipage} \\
\midrule()
\endhead
\texttt{cubble\_df} &
\texttt{{[}{[}\textless{}-,\ dplyr\_col\_modify,\ key\_data,\ key\_vars,\ key,\ print} \\
\texttt{spatial\_cubble\_df} &
\texttt{{[},\ names\textless{}-,\ \ tbl\_sum,\ dplyr\_reconstruct,\ dplyr\_row\_slice,\ face\_spatial,\ face\_temporal,\ unfold,\ arrange,\ rename,\ rowwise,\ group\_by,\ ungroup,\ select,\ spatial,\ summarise,\ unfold,\ update\_cubble} \\
\texttt{temporal\_cubble\_df} &
\texttt{{[},\ names\textless{}-,\ tbl\_sum,\ arrange,\ dplyr\_reconstruct,\ dplyr\_row\_slice,\ face\_spatial,\ face\_temporal,\ unfold,\ fill\_gaps,\ group\_by,\ ungroup,\ \ rename,\ rowwise,\ scan\_gaps,\ select,\ spatial,\ summarise,\ tbl\_sum,\ bind\_rows,\ bind\_cols,\ update\_cubble} \\
\bottomrule()
\end{longtable}

Specifically, the pair of verbs, \code{face_temporal()} and
\code{face_spatial()}, pivot the cubble object between the spatial and
temporal face of the multivariate spatio-temporal cube, as illustrated
in Figure \ref{fig:face}. The code below uses the function
\code{face_temporal()} and \code{face_spatial()} to transform between a
spatial cubble (\code{cb_nested}) and a temporal one (\code{cb_long}),
introduced earlier in subsection @ref(object):

@@

\caption{An illustration of the function \code{face\_temporal()} and \code{face\_spatial()}: \code{face\_temporal()} converts a spatial cubble (neseted form) into a temporal cubble (long form) to focus on the temporal variables. Conversely, \code{face\_spatial()} transforms a temporal cubble into a spatial one to emphasize spatial variables.}(\#fig:face)
\end{figure}
\end{CodeChunk}

@@

An existing \code{cubble} object can promote its temporal component to a
\code{tsibble} object by applying \code{make_temporal_tsibble()}. The
following code illustrates this with the object \code{cb_long} created
in section @ref(create) and the promoted cubble object is equal to the
cubble object originally created from a \code{tsibble} object:

@@

Similarly, an \code{sf} object can be supplied as the spatial component
to create a \code{cubble} object, with the \code{coords} argument being
omitted. This opens up the possibility to represent fixed area with
polygons or multipolygons (see Applications @ref(covid)) and the
\code{coords} argument will be calculated as the centroids of the
(multi)polygons. The \code{[sf]} print in the cubble header suggest an
spatial component being also a \code{sf} object:

@@

The \code{sf} functions applicable to a \code{cubble} object have been
listed in Table @ref(tab:funs) and the following code shows how to
perform coordinate transformation with \code{st_transform} on a cubble
object:

@@

\caption{Linking between multiple plots. The line plots and the map are constructed from shared \code{crosstalk} objects. When a station is selected on the map (a), the corresponding row in the spatial \code{cubble} will be activated. This will link to all the rows with the same id in the temporal \code{cubble} (b) and update the line plot (c).}(\#fig:illu-interactive)
\end{figure}
\end{CodeChunk}

@@

A cubble object can be created from separate spatial and temporal
component using the \code{make_cubble()} function, introduced in Section
@ref(create). The \code{by} argument is used to specify the joining
variable from the two component using the \code{*_join()} \code{by}
syntax:

@@

\caption[A glyph map of the monthly maximum average temperature for weather stations in  Victoria and New South Wales (NSW) for the periods (1971-1975, 2016-2020)]{A glyph map of the monthly maximum average temperature for weather stations in  Victoria and New South Wales (NSW) for the periods (1971-1975, 2016-2020). The corresponding average time series for the cobar station are display on the top left corner. From the glyph map we can observe that the monthly trend is similar for all locations (low in the winter, high in the summer), and small increased temperatures, particularly in late summer can be seen at most stations in NSW.}(\#fig:basic-manip)
\end{figure}
\end{CodeChunk}

@@

\caption[Weather stations and river gauges with matched pairs labelled on the map (a) and plotted across time (b)]{Weather stations and river gauges with matched pairs labelled on the map (a) and plotted across time (b). Precipitation and water level have been standardised between 0 and 1 to be displayed on the same scale. The water level reflects the increase in precipitation. The numbers (1, 7, 6, 9) indicate the group index derived from spatial matching, only those that were selectd by temporal matching are shown here.}(\#fig:matching)
\end{figure}
\end{CodeChunk}

@@

\caption[A reproduction of the second row (ERA5 data) of Figure 19 in Hersbach et al (2020) to illustrate the break-up of sourthern polar vortex in late September and early October 2002]{A reproduction of the second row (ERA5 data) of Figure 19 in Hersbach et al (2020) to illustrate the break-up of sourthern polar vortex in late September and early October 2002. The polar vortex, signalled by the high specific humidity, splits into two on 2002-09-26 and further splits into four on 2002-10-04.}(\#fig:netcdf)
\end{figure}
\end{CodeChunk}

@@

\caption[Exploring temperature variation using linking of a map and seasonal display]{Exploring temperature variation using linking of a map and seasonal display. Each row is a screen dump of the process. The top row shows all locations and all temperature profiles. Selecting a particular location on the map (here Mount Elizabeth) produces the plot in the second row. The maximum and minimum temperatures are shown using a ribbon. The bottom row first selects the lowest temperature in August in the seasonal display, which highlights the corresponding station on the map (Thredbo). Another  station, located in the Tasmania Island, is then selected to compare its temperature variation with the Thredbo station.}(\#fig:interactive-linking)
\end{figure}
\end{CodeChunk}

@@

This paper presents an \proglang{R} package \pkg{cubble} for organizing,
manipulating and visualizing spatio-temporal data. The package
introduces a new data class for spatio-temporal data, \code{cubble},
that connects the time invariant and varying variables and that allows
the user to work with a nested and long form of the data. This work adds
capabilities into the spatio-temporal practitioners toolbox to integrate
it with a tidy data framework. The data structure and functions
introduced in this package can be used and combined with existing
spatial data analysis packages such as \pkg{sf}, data wrangling packages
such as \pkg{dplyr}, and visualization packages such as \pkg{ggplot2},
\pkg{plotly} and \pkg{leaflet}.

Numerous examples are provided in the main text, appendix and package
vignettes. These include creating and coercing data with mismatched
names, handling hierarchical data, matching time series spatially and
temporally, reproducing ERA5 plots from NetCDF data. Visualization of
the \code{cubble} objects can be done with interactive graphic pipelines
using \pkg{crosstalk}, \pkg{plotly} and \pkg{leaflet}.

There are several possible future directions for the work. The data
structure only described fixed spatial sites, and it could be useful to
provide tools to accommodate moving coordinates as might be encountered
in animal movement data. That could be achieved with a list-column for
the location coordinates, and an additional form that these locations
can be pivoted into, like the long form for temporal variables. For
multiple measured variables, the \pkg{cubble} package could be
integrated with dimension reduction methods, and dynamic graphics for
multiple dimensions such as a tour \citep{wickham_tourr_2011}.

>>>>

\author{
H. Sherry Zhang\\Monash University \And Dianne Cook\\Monash University \And Ursula Laa\\University of Natural \\ Resources and Life Sciences \AND Nicolas Langrené\\BNU-HKBU \\ United International College \And Patricia Menéndez\\Monash University
}
\title{\pkg{cubble}: An \proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data}

@@

      Ursula Laa\\
    University of Natural \\ Resources and Life Sciences\\
    Gregor-Mendel-Straße 33, 1180 Wien, Austria\\
  E-mail: \href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}\\

      Nicolas Langrené\\
    BNU-HKBU \\ United International College\\
    2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China\\
  E-mail: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}\\

@@

The rest of the paper is organized as follows: Section \ref{cubble} presents the main design and functionality of the \pkg{cubble} package. Section \ref{others} explains how the \pkg{cubble} package deals with more advanced considerations, including data with hierarchical structure, data matching and how the package fits with existing static and interactive visualization tools. Moreover we also illustrate how the \pkg{cubble} package deals with spatio-temporal data transformations. Section \ref{examples} uses Australian weather station data and river level data as examples to demonstrate the use of the package. An example of how the \pkg{cubble} package handles Network Common Data Form (NetCDF) data is also provided. Section \ref{conclude} discuss the paper contributions and future directions.

@@

\caption[The full space-time and sparse space-time layouts handled by the cubble package]{The full space-time and sparse space-time layouts handled by the cubble package. Reproduced from Figure 1 in the spacetime paper \citep{spacetime}.}\label{fig:layout}
\end{figure}
\end{CodeChunk}

@@

In a spatial cubble, spaital variables are organised as columns and temporal variables are nested within a specialised \texttt{ts} column. The cubble object printed below contains weather records in three airport stations (the creation of a cubble object will be explained in Section \ref{create}). This toy data is a subset of a larger data \code{climate_aus} collected from Global Historical Climatology Network Daily (GHCND). It records three airport stations located in Melbourne, Australia and includes spatial variables such as station ID, longitude, latitude, elevation, station name, World Meteorology Organisation ID. The dataset contains temporal variables including precipitation, maximum and minimum temperature, which can be read from the cubble header:

@@

The function \code{make_cubble()} composes a \code{cubble} object from a spatial table (\code{spatial}) and a temporal table (\code{temporal}), along with three attributes introduced in the subsection \ref{object}: \code{key}, \code{index}, and \code{coords}. The following code creates the spatial \code{cubble} object shown previously:

@@

Table \ref{tab:funs} summarises the functions implemented in the \pkg{cubble} package and Table \ref{tab:methods} details the methods implemented for each of the three cubble classes. The \code{cubble_df} class handles methods that behave consistently in both spatial (\code{spatial_cubble_df}) and temporal cubble (\code{temporal_cubble_df}). When the method has distinct behavior for the spatial cubble and temporal cubble, it is implemented separately for each subclass.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8750}}@{}}
\caption{\label{tab:funs} Functions implemented in the \pkg{cubble} package}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Category
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Functions
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Category
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Functions
\end{minipage} \\
\midrule()
\endhead
base R & \texttt{{[}}, \texttt{{[}{[}\textless{}-}, \texttt{names\textless{}-} \\
tidyverse & \texttt{dplyr\_row\_slice}, \texttt{dplyr\_col\_modify}, \texttt{dplyr\_reconstruct}, \texttt{select}, \texttt{mutate}, \texttt{arrange}, \texttt{filter}, \texttt{group\_by}, \texttt{ungroup}, \texttt{summarise}, \texttt{select}, \texttt{slice}, \texttt{rowwise}, \texttt{rename}, \texttt{bind\_rows}, \texttt{bind\_cols}, \texttt{relocate}, \texttt{type\_sum}, the slice family (\texttt{slice\_head}, \texttt{slice\_tail}, \texttt{slice\_max}, \texttt{slice\_min}, \texttt{slice\_sample}) and the join family (\texttt{left\_join}, \texttt{right\_join}, \texttt{inner\_join}, \texttt{full\_join}, \texttt{anti\_join}, \texttt{semi\_join}) \\
cubble & \texttt{as\_cubble}, \texttt{cubble}, \texttt{make\_cubble}, \texttt{check\_key} \texttt{face\_temporal}, \texttt{face\_spatial}, \texttt{unfold}, \texttt{key}, \texttt{key\_vars}, \texttt{key\_data}, \texttt{index}, \texttt{index\_var}, \texttt{coords}, \texttt{spatial}, \texttt{match\_sites}, \texttt{match\_spatial}, \texttt{match\_temporal}, \texttt{geom\_glyph}, \texttt{geom\_glyph\_box}, \texttt{geom\_glyph\_line}, \texttt{make\_spatial\_sf}, \texttt{make\_temporal\_tsibble}, \texttt{fill\_gaps}, and \texttt{scan\_gaps} \\
\bottomrule()
\end{longtable}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7500}}@{}}
\caption{\label{tab:methods} Methods implemented for the three \code{cubble} classes.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Class
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Methods
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Class
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Methods
\end{minipage} \\
\midrule()
\endhead
\texttt{cubble\_df} & \texttt{{[}{[}\textless{}-,\ dplyr\_col\_modify,\ key\_data,\ key\_vars,\ key,\ print} \\
\texttt{spatial\_cubble\_df} & \texttt{{[},\ names\textless{}-,\ \ tbl\_sum,\ dplyr\_reconstruct,\ dplyr\_row\_slice,\ face\_spatial,\ face\_temporal,\ unfold,\ arrange,\ rename,\ rowwise,\ group\_by,\ ungroup,\ select,\ spatial,\ summarise,\ unfold,\ update\_cubble} \\
\texttt{temporal\_cubble\_df} & \texttt{{[},\ names\textless{}-,\ tbl\_sum,\ arrange,\ dplyr\_reconstruct,\ dplyr\_row\_slice,\ face\_spatial,\ face\_temporal,\ unfold,\ fill\_gaps,\ group\_by,\ ungroup,\ \ rename,\ rowwise,\ scan\_gaps,\ select,\ spatial,\ summarise,\ tbl\_sum,\ bind\_rows,\ bind\_cols,\ update\_cubble} \\
\bottomrule()
\end{longtable}

Specifically, the pair of verbs, \code{face_temporal()} and \code{face_spatial()}, pivot the cubble object between the spatial and temporal face of the multivariate spatio-temporal cube, as illustrated in Figure \ref{fig:face}. The code below uses the function \code{face_temporal()} and \code{face_spatial()} to transform between a spatial cubble (\code{cb_nested}) and a temporal one (\code{cb_long}), introduced earlier in subsection \ref{object}:

@@

\caption{An illustration of the function \code{face\_temporal()} and \code{face\_spatial()}: \code{face\_temporal()} converts a spatial cubble (neseted form) into a temporal cubble (long form) to focus on the temporal variables. Conversely, \code{face\_spatial()} transforms a temporal cubble into a spatial one to emphasize spatial variables.}\label{fig:face}
\end{figure}
\end{CodeChunk}

@@

An existing \code{cubble} object can promote its temporal component to a \code{tsibble} object by applying \code{make_temporal_tsibble()}. The following code illustrates this with the object \code{cb_long} created in section \ref{create} and the promoted cubble object is equal to the cubble object originally created from a \code{tsibble} object:

@@

Similarly, an \code{sf} object can be supplied as the spatial component to create a \code{cubble} object, with the \code{coords} argument being omitted. This opens up the possibility to represent fixed area with polygons or multipolygons (see Applications \ref{covid}) and the \code{coords} argument will be calculated as the centroids of the (multi)polygons. The \code{[sf]} print in the cubble header suggest an spatial component being also a \code{sf} object:

@@

The \code{sf} functions applicable to a \code{cubble} object have been listed in Table \ref{tab:funs} and the following code shows how to perform coordinate transformation with \code{st_transform} on a cubble object:

@@

\caption{Linking between multiple plots. The line plots and the map are constructed from shared \code{crosstalk} objects. When a station is selected on the map (a), the corresponding row in the spatial \code{cubble} will be activated. This will link to all the rows with the same id in the temporal \code{cubble} (b) and update the line plot (c).}\label{fig:illu-interactive}
\end{figure}
\end{CodeChunk}

@@

A cubble object can be created from separate spatial and temporal component using the \code{make_cubble()} function, introduced in Section \ref{create}. The \code{by} argument
is used to specify the joining variable from the two component using the \code{*_join()} \code{by} syntax:

@@

\caption[A glyph map of the monthly maximum average temperature for weather stations in  Victoria and New South Wales (NSW) for the periods (1971-1975, 2016-2020)]{A glyph map of the monthly maximum average temperature for weather stations in  Victoria and New South Wales (NSW) for the periods (1971-1975, 2016-2020). The corresponding average time series for the cobar station are display on the top left corner. From the glyph map we can observe that the monthly trend is similar for all locations (low in the winter, high in the summer), and small increased temperatures, particularly in late summer can be seen at most stations in NSW.}\label{fig:basic-manip}
\end{figure}
\end{CodeChunk}

@@

\caption[Weather stations and river gauges with matched pairs labelled on the map (a) and plotted across time (b)]{Weather stations and river gauges with matched pairs labelled on the map (a) and plotted across time (b). Precipitation and water level have been standardised between 0 and 1 to be displayed on the same scale. The water level reflects the increase in precipitation. The numbers (1, 7, 6, 9) indicate the group index derived from spatial matching, only those that were selectd by temporal matching are shown here.}\label{fig:matching}
\end{figure}
\end{CodeChunk}

@@

\caption[A reproduction of the second row (ERA5 data) of Figure 19 in Hersbach et al (2020) to illustrate the break-up of sourthern polar vortex in late September and early October 2002]{A reproduction of the second row (ERA5 data) of Figure 19 in Hersbach et al (2020) to illustrate the break-up of sourthern polar vortex in late September and early October 2002. The polar vortex, signalled by the high specific humidity, splits into two on 2002-09-26 and further splits into four on 2002-10-04.}\label{fig:netcdf}
\end{figure}
\end{CodeChunk}

@@

\caption[Exploring temperature variation using linking of a map and seasonal display]{Exploring temperature variation using linking of a map and seasonal display. Each row is a screen dump of the process. The top row shows all locations and all temperature profiles. Selecting a particular location on the map (here Mount Elizabeth) produces the plot in the second row. The maximum and minimum temperatures are shown using a ribbon. The bottom row first selects the lowest temperature in August in the seasonal display, which highlights the corresponding station on the map (Thredbo). Another  station, located in the Tasmania Island, is then selected to compare its temperature variation with the Thredbo station.}\label{fig:interactive-linking}
\end{figure}
\end{CodeChunk}

@@

This paper presents the \proglang{R} package \pkg{cubble} for organizing, wrangling and visualizing spatio-temporal data. The package introduces a new data class, \code{cubble}, consisting of two subclass, spatial cubble and a temporal cubble, to organise spatio-temporal data in two different formats within the tidy data framework. The data structure and functions introduced in this package can be used and combined with existing spatial and temporal data analysis packages such as \pkg{sf} and \pkg{tsibble}, data wrangling packages such as \pkg{dplyr}, and visualization packages such as \pkg{ggplot2}, \pkg{plotly}, and \pkg{leaflet}.

The paper includes numerous examples to illustrate the utility of \pkg{cubble} as a data structure for spatio-temporal analysis. These examples cover different tasks of a typical analysis workflow: handling data with spatial and temporal misalignment, matching data from multiple sources, and creating both static and interactive spatio-temporal visualisation. For future directions, other commonly-used spatial or temporal data structures can be integrated into \pkg{cubble} to extend analysts' familiar spatial and temporal toolkit to spatio-temporal.
